Cross-scale integration of knowledge for predicting species ranges: a metamodeling framework

### 2015-07-31 tex/appendix_s1.tex
Added nicer plot map to appendix
Additional revisions for Reviewer 1

\begin{figure}[t]
\includegraphics{figs/plotmap.pdf}
\caption{Map of permanent forest plot locations}
\label{fig:plotmap}
\end{figure}

We divided the dataset into calibration and validation subsets, with 1/3 of the data reserved for validation and the remaining 2/3 used for calibration.
As in the previous example, we modeled the probability of presence as a function of climate.
We considered seven climate variables: the number of degree days (ddeg), minimum temperature (min\_temp), potential evapotranspiration (PET), annual precipitation (an\_prcp), summer precipitation (sum\_prcp), winter precipitation (win\_prcp), and the ratio of annual precipitation to potential evapotranspiration (pToPET).
See \citet{Morin2009} for a complete description of these variables.
For projection, we used predictions for \rev{2100} from the HadCM3 GCM \citep{Pope2000} driven by the A2 emission scenario \citep{Nakicenovic2000}, and used the parameters of the models to forecast suitability into the future.
Exploratory analysis revealed that there were collinearities within the predictor variables.
Thus, we only included three variables in the model: ddeg, an\_prcp, and pToPET.
To select the form of the naive model (and thus the metamodel), we used stepwise regression with BIC as the evaluation criteria.
The search scope included all models with linear, quadratic, and cubic terms for all three variables.
Because Phenofit provided predictions in the form of a probability of presence (rather than directly in terms of the metamodel parameters as in the previous example), it was necessary to develop a method for relating these predictions to the metamodel.
We opted to do this via simulation because it is a natural method to use in an MCMC scheme.
For each MCMC iteration, we generated a simulated dataset by drawing \(N_x\) samples with probability \(p_x\) from a Binomial distribution, where \(N_x\) is the number of observations in the occurrence dataset for a given grid cell \(x\), and \(p_x\) is the predicted probability of presence from Phenofit for that cell.
The metamodel was then fit to these simulated data (with prior conditioning on the naive model).
This procedure, by generating a random dataset with each iteration, incorporated the variance in the Phenofit predictions.
Convergence for all models was assessed by comparing the results of multiple chains with overdispersed starting points.
Based on the results of these initial runs, we thinned all posterior samples by recording only every 50th sample, and we obtained 25000 posterior samples for each model following a burn-in period of 20000 samples.

>>>>

\begin{figure}[t]
\includegraphics[width=4.5in]{figs/plotmap.png}
\caption{Map of permanent forest plot locations}
\label{fig:plotmap}
\end{figure}

We divided the dataset into calibration and validation subsets, with 1/3 of the data reserved for validation and the remaining 2/3 used for calibration.
As in the previous example, we modeled the probability of presence as a function of climate.
We considered seven climate variables: the number of degree days (ddeg), minimum temperature (min\_temp), potential evapotranspiration (PET), annual precipitation (an\_prcp), summer precipitation (sum\_prcp), winter precipitation (win\_prcp), and the ratio of annual precipitation to potential evapotranspiration (pToPET).
See \citet{Morin2009} for a complete description of these variables.
For projection, we used predictions for \rev{2100} from the HadCM3 GCM \citep{Pope2000} driven by the A2 emission scenario \citep{Nakicenovic2000}, and used the parameters of the models to forecast suitability into the future.
Exploratory analysis revealed that there were collinearities within the predictor variables.
Thus, we only included three variables in the model: ddeg, an\_prcp, and pToPET.
To select the form of the naive model (and thus the metamodel), we used stepwise regression with BIC as the evaluation criteria.
The search scope included all models with linear, quadratic, and cubic terms for all three variables.
\rev{Thus, the naive model (which also defined the scope of the metamodel), can be formulated as a simple Binomial GLM, as in Example 1:
\begin{equation}
\psi_N =\text{logit}^{-1}\left( \theta_M D_M \right)
\end{equation}
where \(\psi_N\) is the probability of presence in the naive model, \(\theta_M\) is the parameter vector of the metamodel (presently unconstrained by any additional information), and \(D_M\) is the macroclimate covariate matrix at the metamodel scale.
The parameter vector is estimated following equation 2 in the main text:
\begin{equation}
p\left (\theta_M \mid X_M,D_M \right ) \propto 
p \left(X_M \mid \theta_M, D_M \right)
p \left(\theta_M \right)
\end{equation}

Because Phenofit provided predictions in the form of a probability of presence (rather than directly in terms of the metamodel parameters as in the previous example), it was necessary to relate these predictions to the metamodel.
We opted to do this via simulation because it is a natural method to use in an MCMC scheme.
We thus treated the Phenofit predictions (\(\psi_P\)) as an observed result of a true underlying set of occurrences (\(X_P\)).
Because these occurrences were unobserved (and in the case of the future distribution, unobservable), it was necessary to infer them using MCMC.
For each MCMC iteration, we generated a simulated occurrence dataset \(\hat{X}_P\) by drawing from a Binomial distribution with probabilities equal to \(\psi_P\).
In order to give both the occurrence dataset and Phenofit predictions equal weight, we drew the same number of samples as were present in the occurrence dataset.
The metamodel was then fit to these simulated data (with conditioning on the naive model as well):
\begin{equation} 
\overbrace{p(\theta_M \mid X_M, D_M, X_P, \psi_P)}^\text{integrated posterior}
\propto
\overbrace{p\left (X_P \mid \theta_M,D_M,\psi_P \right )}^{\substack{\text{new information} \\ \text{from Phenofit}}}
\overbrace{p \left(X_M \mid \theta_M, D_M \right) p \left(\theta_M \right)}^{\substack{\text{naive metamodel} \\ \text{posterior}}}
\overbrace{p \left(\psi_P \right)}^{\substack{\text{prior for} \\ \text{Phenofit}}}	
\end{equation}
This procedure, by generating a random dataset with each iteration, incorporated the variance in the Phenofit predictions.
Finally, because we were interested in two separate questions with respect to Phenofit integration (one regarding how integration affects uncertainty in our present estimate of the species distribution and a one regarding the effect of integration on projections into the future), we performed two separate integrations with the Phenofit information.
To address the first question, we used the Phenofit predictions for the present as \(\psi_P\) and present climate as \(D_M\), producing the Integrated-Present model.
To address the second, we used the future predictions from Phenofit and future climate, resulting in the Integrated-Future model.
In both cases, the naive model was a GLM relating present occurrences to present climate (and thus the prior response to the environment was static, similarly to how one might use the same conventional SDM to both model present distribution and project into the future).
Convergence for all models was assessed by comparing the results of multiple chains with overdispersed starting points.
Based on the results of these initial runs, we thinned all posterior samples by recording only every 50th sample, and we obtained 25000 posterior samples for each model following a burn-in period of 20000 samples.}

\section*{Implementation}
\subsection*{Model weights}
\rev{As a start, users may consider weighting models based on sample size or number of independent treatments, as we have done in example 1, where the weight for the sub-model of 0.05 was chosen based on the 5 precipitation levels relative to 100 independent data points in the correlative dataset.
Because we are integrating model predictions, a highly precise model model with a small sample size (e.g., an experiment with a small number of treatment levels but strong effects) may provide too strong of a constraint on the metamodel.}

### 2015-07-31 tex/manuscript.tex
Additional revisions for Reviewer 1

	\textbf{Main text word count: } 4595 %WORD_COUNT %Target 5000

@@

We accomplish model integration by treating the \(\psi_S\), the posterior predictions of the submodel\(\theta_S\), as prior information about some of the parameters of \(\theta_M\) (i.e., parameters related to precipitation), expanding Equation \ref{eq:ex1_bayes} to incorporate the new information from the sub-model:
%-----
\begin{equation}
	\label{eq:ex1_integrated}
	\overbrace{p(\theta_M \mid X_M, T_M, P_M, \theta_S, \psi_S)}^\text{integrated posterior}
	\propto
	\overbrace{p\left (\psi_S \mid \theta_M,P_M \right )}^{\substack{\text{new information} \\ \text{from sub-model}}}
	\overbrace{p \left(X_M \mid \theta_M, T_M, P_M \right) p \left(\theta_M \right)}^{\substack{\text{naive metamodel} \\ \text{posterior}}}
	\overbrace{p \left(\theta_S \right)}^{\substack{\text{prior for} \\ \text{sub-model}}}	
\end{equation}
%-----
%\dg{There is something wrong with the following equation, or I missed the reasonning somewhere in the last year. In any case, I tried to understand the logic from the text and haven't succeed at it. Either it needs to be fixed, or it needs to be better explained. If I don't get it, my prior tells me nobody will. This equation is central to the paper now that the maths have been put in the appendix. I was expecting to see the likelihood of X\_M given the parameters theta\_M.  The prior of theta\_M being the posterior of the model theta\_M applied to the data psi\_S. The actual likelihood gives the probability of observing the prediction of the sub-model given the parameters of the sub-model. There is not integration on the conditions, it only results from the product of the probabilities.}
As before, the metamodel \(\theta_M\) can be used to predict probability of occurrence \((\psi_I)\).
However, these predictions now reflect the presence-absence data \(X_M\) as well as the information from \(\theta_S\), including all of the data sources used to produce this sub-model.
Finally, we note the presence of marginal distributions for both models (i.e., \(p(\theta_M)\) and \(p(\theta_S)\)).
These can be informative (e.g., incorporating further prior information or the predictions of additional sub-models), semi-informative (e.g., to provide greater weight to more informative models), or uninformative.
For purposes of this example, we applied prior weights of 20 and 1 to the correlative and mechanistic models, respectively, reflecting the increased generality and much larger sample size of the correlative data.
This procedure has the effect of increasing the variance of the model and prevents biasing the parameter estimation in favor of the mechanistic model.

@@

\subsection*{Example 2: Constraining an SDM using phenological information}
For the second example, we consider the problem of forecasting the potential distribution of a species following climate change.
There is considerable interest in comparing correlative and mechanistic projections with respect to climate change \citep{Morin2009}, and correctly characterizing uncertainty is a critical aspect of this problem \citep{Cheaib2012}.
Despite being a relatively common application of SDMs \citep{Guisan2005}, projecting models parameterized with modern climate data to future climate scenarios remains problematic \citep{Araujo2006}.
We used our framework to constrain a climate-based SDM with information obtained from Phenofit, a mechanistic model that predicts a species' probability of presence as a function of the suitability of the environment given the species' phenology \citep{Chuine2001, Morin2009}.
Here we describe briefly the dataset, methods, and the results of the analysis (see \rev{Appendix S2} for data and scripts to reproduce the analysis).

@@

To perform the integration, we constrained the estimates of the naive model with the additional information from Phenofit while considering two different potential modeling goals.
The first is improving our model of the present range of the species.
Using both datasets to develop a range model for the species has a number of advantages.
Assuming we have chosen climate variables that well represent the constraints on the species, including Phenofit in our model is likely to reduce bias in our estimate of the fundamental niche.
The posterior predictions of the model (that is, the probability of presence projected into geographic space) will incorporate uncertainty from all sources, including the modeling process, data, parameterization, and the ecological processes included.
This can provide a much more accurate estimate of the uncertainty of our predictions.
Thus, for our first integration we combined the naive model with the Phenofit predictions for the present; we refer to this model as ``Integrated-Present'' hereafter.
We also validated this model and compared it against the naive model by computing the area under the receiver operating curve (AUC), a commonly-used metric that measures the rate of correct classification and varies from 0 to 1 (where a value of 1 indicates perfect classification and 0.5 indicates no difference from a random model) \citep{Swets1988}.
The second modeling goal is projecting changes to the range of sugar maple following climate change.
Process-based models and correlative models often differ substantially when projecting beyond the range of the original data (see Example 1).
Thus, we obtained predictions for 2100 from Phenofit \citep{Morin2009} and used them to condition the metamodel given the results of the naive model under future climate.
This procedure provided a consensus view of the future range of the species; we refer to this model as ``Integrated-Future.''

@@

The Integrated-Future model presents a different interpretation of the response of sugar maple to warmer temperatures. 
The naive model predicted a substantial northward migration; in other words, the expectation under the naive model is that the present occurrence data is an accurate estimation of the niche, and thus the species should track its niche northward as the climate warms.
The Integrated-Future model, in contrast, predicted substantially more tolerance to warmer temperatures, reflecting similar predictions from Phenofit (Figs. \ref{fig:ex2_response}, \ref{fig:ex2_map_fut}).
Although both models predicted a northward range shift, the change under Phenofit was limited to approximately 200 km north of the present range limit of the species, compared with more than 900 km for the naive model.
Phenofit also predicted little change in the southern range limit of the species, while the naive model projected loss of the species over much of the southern portion of the range (Fig. \ref{fig:ex2_map_fut}).

@@

\subsection*{Comparison with other methods}
The methods provided here expand upon the motivation of hybrid models to develop a more robust approach for using ecological models for prediction while overcoming some limitations characteristic of other integrated approaches.
In particular, it is often difficult in a hybrid model to identify parameters that can be used to connect different modeling frameworks and produce a meaningful response \citep{Thuiller2013}. 
The difficulty of including information from experimental studies or ecological processes at lower scales can be a possible drawback of hybrid models \citep{Smolik2010, Thuiller2014a}. 
Bayesian methods such as our are able to overcome some of these limitations, as they provide a natural framework for the incorporation of multiple sources of information.
Because the metamodel is expressed as a series of conditional probabilities (see Appendix S1), this information can be included simply as long as the probability of the metamodel can be expressed mathematically.
Furthermore, Bayesian methods produce posterior distributions of parameters and predictions, rather than point estimates, allowing for a comprehensive understanding of the uncertainty. 
Finally, Bayesian methods inherently allow for feedbacks or interactions between sub models, which may be a more realistic representation of ecological dynamics where many factors may simultaneously influence the system.

@@

Determining the functions to use to express the likelihood of the sub-models given the metamodel (i.e., Eq. \ref{eq:ex1_integrated}) is a critical point.
The challenge is three fold: (1) which spatial and temporal scales (i.e. which processes), are to be considered, (2) how to build a scale the sub-models to be consistent with the metamodels, and (3) how error and uncertainty propagate from the sub-models to the metamodel. 
Although we argue that our proposed framework is able to easily deal with very different scales and that the Bayesian framework allows for an efficient integration of uncertainty throughout all scales considered, the building of scaling functions is an object of investigation on its own. 
It is likely that the modeling process may include multiple functions operating at different scales when taking all known processes and models into account. 
Indeed, if species distributions are a function of, e.g. population growth rate \citep{Guisan2000}, they will involve processes at the individual (e.g. competition) or cellular (e.g. photosynthesis) scales. Such very large differences in spatial scales would require more sophisticated upscaling methods than simple functions such as those we have used here. Our framework is still applicable whatever the chosen upscaling approach and is able to propagate uncertainties from sub-models to the metamodel.

>>>>

	\textbf{Main text word count: } 5083 %WORD_COUNT %Target 5000

@@

We accomplish model integration by treating \(\psi_S\), the posterior predictions of the submodel \(\theta_S\), as prior information about some of the parameters of \(\theta_M\) (i.e., parameters related to precipitation), expanding Equation \ref{eq:ex1_bayes} to incorporate the new information from the sub-model:
%-----
\begin{equation}
	\label{eq:ex1_integrated}
	\overbrace{p(\theta_M \mid X_M, T_M, P_M, \theta_S, \psi_S)}^\text{integrated posterior}
	\propto
	\overbrace{p\left (\psi_S \mid \theta_M,P_M \right )}^{\substack{\text{new information} \\ \text{from sub-model}}}
	\overbrace{p \left(X_M \mid \theta_M, T_M, P_M \right) p \left(\theta_M \right)}^{\substack{\text{naive metamodel} \\ \text{posterior}}}
	\overbrace{p \left(\theta_S \right)}^{\substack{\text{prior for} \\ \text{sub-model}}}	
\end{equation}
%-----
%\dg{There is something wrong with the following equation, or I missed the reasonning somewhere in the last year. In any case, I tried to understand the logic from the text and haven't succeed at it. Either it needs to be fixed, or it needs to be better explained. If I don't get it, my prior tells me nobody will. This equation is central to the paper now that the maths have been put in the appendix. I was expecting to see the likelihood of X\_M given the parameters theta\_M.  The prior of theta\_M being the posterior of the model theta\_M applied to the data psi\_S. The actual likelihood gives the probability of observing the prediction of the sub-model given the parameters of the sub-model. There is not integration on the conditions, it only results from the product of the probabilities.}
As before, the metamodel \(\theta_M\) can be used to predict probability of occurrence \((\psi_I)\).
However, these predictions now reflect the presence-absence data \(X_M\) as well as the information from \(\theta_S\), including all of the data sources used to produce this sub-model.
Finally, we note the presence of marginal distributions for both models (i.e., \(p(\theta_M)\) and \(p(\theta_S)\)).
These can be informative (e.g., incorporating further prior information or the predictions of additional sub-models), semi-informative (e.g., to provide greater weight to more informative models), or uninformative.
For purposes of this example, we applied prior weights of 1 and 0.05 to the correlative and mechanistic models, respectively, reflecting the increased generality and much larger sample size of the correlative data.
This procedure has the effect of increasing the variance of the model and prevents biasing the parameter estimation in favor of the mechanistic model.

@@

\subsection*{Example 2: Constraining an SDM using phenological information}
For the second example, we consider the problem of forecasting the potential distribution of a species following climate change.
There is considerable interest in comparing correlative and mechanistic projections with respect to climate change \citep{Morin2009}, and correctly characterizing uncertainty is a critical aspect of this problem \citep{Cheaib2012}.
Despite being a relatively common application of SDMs \citep{Guisan2005}, projecting models parameterized with modern climate data to future climate scenarios remains problematic \citep{Araujo2006}.
We used our framework to constrain a climate-based SDM with information obtained from Phenofit, a mechanistic model that predicts a species' probability of presence as a function of the suitability of the environment given the species' phenology \citep{Chuine2001, Morin2009}.
Here we describe briefly the dataset, methods, and the results of the analysis (see \rev{ Appendix S1 for implementation details and Appendix S2} for data and scripts to reproduce the analysis).

@@

To perform the integration, we constrained the estimates of the naive model with the additional information from Phenofit while considering two different potential modeling goals.
The first is improving our model of the present range of the species.
Using both datasets to develop a range model for the species has a number of advantages.
Assuming we have chosen climate variables that well represent the constraints on the species, including Phenofit in our model is likely to reduce bias in our estimate of the fundamental niche.
The posterior predictions of the model (that is, the probability of presence projected into geographic space) will incorporate uncertainty from all sources, including the modeling process, data, parameterization, and the ecological processes included.
This can provide a much more accurate estimate of the uncertainty of our predictions.
Thus, for our first integration we combined the naive model with the Phenofit predictions for the present; we refer to this model as ``Integrated-Present'' hereafter.
We also validated this model and compared it against the naive model by computing the area under the receiver operating curve (AUC), a commonly-used metric that measures the rate of correct classification and varies from 0 to 1 (where a value of 1 indicates perfect classification and 0.5 indicates no difference from a random model) \citep{Swets1988}.
The second modeling goal is projecting changes to the range of sugar maple following climate change.
Process-based models and correlative models often differ substantially when projecting beyond the range of the original data (see Example 1).
Thus, we obtained predictions for 2100 from Phenofit \citep{Morin2009} and used them to condition the metamodel given the results of the naive model under future climate.
This procedure provided a consensus view of the future range of the species; we refer to this model as ``Integrated-Future.''
\rev{In both cases, it was necessary to scale the Phenofit predictions, which were probabilistic and not directly derived from the macroclimatic variables used in to fit the naive model, to make them compatible with the naive model, which was based on occurrence data.
We used a latent variable approach, where an unobserved occurrence dataset similar to the one used for the naive model was generated at each MCMC iteration by drawing occurrences from a Binomial distribution with probabilities equal to the probability of occurrence from Phenofit (see Appendix S1 for a full statistical presentation of the model).
This has several advantages.
In addition to expressing the information from Phenofit in a way that is compatible with the naive model, the uncertainty in the predictions is automatically propagated (by simulating a new pseudo-dataset at each iteration), and it becomes possible to address future distribution via simulated future occurrences, (which, by nature, are unobservable).}

@@

The Integrated-Future model presents a different interpretation of the response of sugar maple to warmer temperatures. 
The naive model predicted a substantial northward migration; in other words, the expectation under the naive model \rev{is that the present estimate of the realized niche (obtained using occurrence data) is an unbiased reflection of the fundamental niche; thus the species should track temperature northward as the climate warms}.
The Integrated-Future model, in contrast, predicted substantially more tolerance to warmer temperatures, reflecting similar predictions from Phenofit (Figs. \ref{fig:ex2_response}, \ref{fig:ex2_map_fut}).
\rev{This is because Phenofit estimates different aspects of the realized niche.}
Although both models predicted a northward range shift, the change under Phenofit was limited to approximately 200 km north of the present range limit of the species, compared with more than 900 km for the naive model.
Phenofit also predicted little change in the southern range limit of the species, while the naive model projected loss of the species over much of the southern portion of the range (Fig. \ref{fig:ex2_map_fut}).
\rev{The metamodel thus presents a consensus view of the niche of the species with respect to the macroclimatic variables included in the model (Fig. \ref{fig:ex2_map_pres}), incorporating the present range of the species (using the occurrence data) and information from Phenofit on what conditions will be tolerable in the future.}

@@

\subsection*{Comparison with other methods}
The methods provided here expand upon the motivation of hybrid models to develop a more robust approach for using ecological models for prediction while overcoming some limitations characteristic of other integrated approaches.
In particular, it is often difficult in a hybrid model to identify parameters that can be used to connect different modeling frameworks and produce a meaningful response \citep{Thuiller2013}. 
The difficulty of including information from experimental studies or ecological processes at lower scales can be a possible drawback of hybrid models \citep{Smolik2010, Thuiller2014a}. 
\rev{Bayesian methods provide a natural framework for the incorporation of multiple sources of information, making them an attractive alternative for SDMs.
Hierarchical models in particular have the potential to capture many of the intricacies necessary for implementing hybrid models \citep{Latimer2006}.
\citet{Pagel2012} developed a hybrid approach to species distributions via a dynamic range model.
Similar to our approach, their model integrated demographic information, abundances, and presence/absence data within a hierarchical Bayesian framework to predict species ranges.
However, their approach explicitly links the modelled processes to occurrences/abundances via a detailed demographic model, requiring data that may not be available for many species.
In contrast, our approach allows for the inclusion of more incomplete datasets because integration is performed via the separate predictions of each model (following the application of the scaling function where necessary).}
Because the metamodel is expressed as a series of conditional probabilities (see Appendix S1), this information can be included simply as long as the probability of the metamodel can be expressed mathematically.
Furthermore, Bayesian methods produce posterior distributions of parameters and predictions, rather than point estimates, allowing for a comprehensive understanding of the uncertainty. 
Finally, Bayesian methods inherently allow for feedbacks or interactions between sub models, which may be a more realistic representation of ecological dynamics where many factors may simultaneously influence the system.

@@

Determining the functions to use to express the likelihood of the sub-models given the metamodel (i.e., Eq. \ref{eq:ex1_integrated}) is a critical point.
The challenge is three fold: (1) which spatial and temporal scales (i.e. which processes), are to be considered, (2) how to build a scale the sub-models to be consistent with the metamodels, and (3) how error and uncertainty propagate from the sub-models to the metamodel. 
Although we argue that our proposed framework is able to easily deal with very different scales and that the Bayesian framework allows for an efficient integration of uncertainty throughout all scales considered, the building of scaling functions is an object of investigation on its own. 
It is likely that the modeling process may include multiple functions operating at different scales when taking all known processes and models into account. 
Indeed, if species distributions are a function of, e.g. population growth rate \citep{Guisan2000}, they will involve processes at the individual (e.g. competition) or cellular (e.g. photosynthesis) scales. 
Such very large differences in spatial scales would require more sophisticated upscaling methods than simple functions such as those we have used here. 
Our framework is still applicable whatever the chosen upscaling approach and is able to propagate uncertainties from sub-models to the metamodel.
\rev{Indeed, if a sub-model provides poor information (due, for example, to cross-scale nonlinearities in the response to the environment), the resulting metamodel predictions may be worse than the pre-integration naive model.
In general, we advise users of this framework to carefully consider the scaling in their models with respect to the biology of the organism studied, and to use prior model weights to downweight the scaled models when there is uncertainty about the applicability of a sub-model at the metamodel scale.
For instance, in the first example, we applied a model weight to decrease the influence of the mechanistic model on the metamodel (while still retaining some information contained therein), recognizing that such a simplistic model may not scale well.
Finally, as always when modelling ecological systems, we urge humility in the interpretation of model results and suggest the use of model evaluation and validation tools whenever possible.
We provide additional discussion on the implementation of model weights in Appendix S1.}
